Left-Linear Rewriting in Adhesive Categories

### 2024-07-02 concur/final.tex

\author{Paolo Baldan} 
{Department of Mathematics, University of Padova, Italy}
{baldan@math.unipd.it}
{}
{xxx}

\author{Davide Castelnovo}
{Department of Mathematics, University of Padova, Italy}
{davide.castelnovo@math.unipd.it}
{}
{yyy}

\author{Andrea Corradini}
{Department of Computer Science, University of Pisa, Italy}
 {andrea.corradini@unipi.it}
 {}
 {MUR PRIN 20228KXFN2 ``STENDHAL'' and U. Pisa PRA 2022\_99 ``FM4HD''}

\author{Fabio Gadducci}
{Department of Computer Science, University of Pisa, Italy}
{fabio.gadducci@unipi.it}
{0000-0002-9123-9089}
 {MUR PRIN 20228KXFN2 ``STENDHAL'' and U. Pisa PRA 2022\_99 ``FM4HD''}

@@

  \caption{Rewirting rules for the coffee example.}
  \label{fi:coffee-rules}
\end{figure}

>>>>

\author{Paolo Baldan} 
{Department of Mathematics, University of Padova, Italy}
{baldan@math.unipd.it}
{https://orcid.org/0000-0001-9357-5599}
{xxx}

\author{Davide Castelnovo}
{Department of Mathematics, University of Padova, Italy}
{davide.castelnovo@math.unipd.it}
{https://orcid.org/0000-0002-5926-5615}
{yyy}

\author{Andrea Corradini}
{Department of Computer Science, University of Pisa, Italy}
 {andrea.corradini@unipi.it}
 {https://orcid.org/0000-0001-6123-4175}
 {MUR PRIN 20228KXFN2 ``STENDHAL'' and U. Pisa PRA 2022\_99 ``FM4HD''}

\author{Fabio Gadducci}
{Department of Computer Science, University of Pisa, Italy}
{fabio.gadducci@unipi.it}
{https://orcid.org/0000-0003-0690-3051}
 {MUR PRIN 20228KXFN2 ``STENDHAL'' and U. Pisa PRA 2022\_99 ``FM4HD''}

@@

  \caption{Rewriting rules for the coffee example.}
  \label{fi:coffee-rules}
\end{figure}

### 2024-07-02 concur/submitted/fusione.tex

  \caption{A sequence of rewriting steps in .}
  \label{fi:coffee-rewriting}
\end{figure}

@@

A possible rewriting sequence is depicted in Fig.~\ref{fi:coffee-rewriting}, applying $\rho_m$, $\rho_c$ and $\rho_r$ in sequence, thus producing a coffee with milk, sugar and cocoa, ready for drinking. A first interesting observation is that after applying rule $\rho_m$ to $G_0$, thus producing graph $G_1$, we can still apply rule $\rho_c$, with a non-injective match. Intuitively, the two rules do not interfere. Rather, $\rho_c$ applied after $\rho_m$ will have no effect on the sugar edge (intuitively, because sugar has been already added). Indeed, the rules could be also be applied in reverse order. Should they be considered independent? Moreover, since rule $\rho_r$ just require the presence of sugar, it could be anticipated, and applied after $r_m$. This might lead to think that $\rho_c$ and $\rho_m$ are independent. However, this is not completely clear, since in absence of $\rho_m$, the application of $\rho_c$ would be essential to enable $\rho_r$. ``There is great chaos under heaven, [hence] the situation is excellent!''

>>>>

  \caption{A sequence of rewriting steps.}
  \label{fi:coffee-rewriting}
\end{figure}

@@

A possible rewriting sequence is depicted in Fig.~\ref{fi:coffee-rewriting}, applying $\rho_m$, $\rho_c$ and $\rho_r$ in sequence, thus producing a coffee with milk, sugar and cocoa, ready for drinking. A first interesting observation is that after applying rule $\rho_m$ to $G_0$, thus producing graph $G_1$, we can still apply rule $\rho_c$, with a non-injective match. Intuitively, the two rules do not interfere. Rather, $\rho_c$ applied after $\rho_m$ will have no effect on the sugar edge (intuitively, because sugar has been already added). Indeed, the rules could be also be applied in reverse order. Should they be considered independent? Moreover, since rule $\rho_r$ just require the presence of sugar, it could be anticipated, and applied after $r_m$. This might lead to think that $\rho_c$ and $\rho_m$ are independent. However, this is not completely clear, since in absence of $\rho_m$, the application of $\rho_c$ would be essential to enable $\rho_r$.

--- new version (according to latest mail, rules not yet up to date) ---

Just to get some basic intuition about left-linear rewriting systems, their capabilities and the problems which can arise, assume we want to model a situation in which a coffee comes with a bag of white sugar and a bag of brown sugar. The situation is represented by the graph $G_0$ in Fig.~\ref{fi:coffee-rewriting}: the loop $\mathit{c}$ is the coffee, the two loops $b$ and $w$  represent the white and brown sugar. The corresponding nodes are connected to the coffee node by an $s$-labelled edge to indicate that they can be possibly added to the coffee. The available rules are in Fig.~\ref{fi:coffee-rules}. Rules $\rho_w$ and $\rho_b$ model the addition of white and brown sugar, respectively, to the coffee.
%
Note that adding one kind of sugar, say $x$, to the coffee is realised
by deleting loop $x$ and merging the corresponding node with the
coffee node so that a self-loop $s$ is created.

Once some sugar has been added, the coffee is ready to drink. This is represented by rule $\rho_r$ which check for the presence of sugar, represented by loop $s$, in the coffee, and bring it to the $cr$ state, coffee ready to drink.

A possible rewriting sequence is depicted in Fig.~\ref{fi:coffee-rewriting}, applying $\rho_w$, $\rho_b$ and $\rho_r$ in sequence, thus producing a coffee with both white and brown sugar, ready for drinking. A first interesting observation is that after applying rule $\rho_w$ to $G_0$, thus producing graph $G_1$, we can still apply rule $\rho_b$, with a non-injective match. Intuitively, once the coffee is ready to be drink because some sugar has been added, adding more sugar will not change its state. Indeed, the rules could be also be applied also in reverse order. Should they be considered independent? Moreover, since rule $\rho_r$ just require the presence of sugar, it could be anticipated, and applied after $r_w$. This might lead to think that $\rho_b$ and $\rho_r$ are independent. However, this is not completely clear, since in absence of $\rho_w$, the application of $\rho_b$ would be essential to enable $\rho_r$.
